Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-11-10 paper/paper.tex
some edits suggested by Grammarly

\begin{document}
\maketitle
\begin{abstract}
We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors. We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models. Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine. More importantly, we argue that rather than trying to contrast the roles of theory and "atheoretical" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant predictive work. Instead, we argue that predictive modeling and machine learning are effective tools for theory testing.
\end{abstract}

@@

Although their analysis in the end does not support this point, we do not disagree that theory can be important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with ``theoretical'' modeling, B\&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms. They do not provide much guidance on an exact specification of what variables and which functional form is required. As a result it is both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to ``explain'' a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

To examine the first question, B\&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

\begin{itemize}
\tightlist
\item
  Escalation: a set of ten indicators, drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other.
\item
  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels. These are directed, thus making for four directed dyads, which with four quad categories make sixteen covariates.
\item
  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a total of four covariates.
\item
  CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling \(1,160\) covariates, which are mostly zero for any country in any month.
\item
  Average: unweighted average of the predictions from the four models briefly described above.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Escalation Only: the original basic escalation model with only ICEWS predictors.
\item
  With PITF Predictors: a random forest that also adds the PITF annual, structural variables.
\item
  Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions.
\item
  PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
\item
  PITF Only: a random forest model based only on the annual, structural PITF model predictors.
\end{itemize}

@@

Finally, B\&S used their escalation model to create forecasts for the first half of 2016. In their third analysis, they score the forecasts accuracy using subsequent civil war onset data. This is summarized in B\&S Table 3.

@@

While replicating and analyzing B\&S's results, we found several issues worthy of further discussion and investigation. These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.\footnote{We also note there are additional concerns arising from the question of how the random forest models were tuned by B\&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.}

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model \textit{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B\&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. B\&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90\%--17 of 19--clearly used empirical ROC curves, and 2 (10\%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey. Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard. What is surprising is that anyone would choose the smoothed ROC implementation.

@@

We believe that B\&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.

As a result: (1) in the test set, all countries with names that alphabetically start after ``Tongo'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical. For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions for that time period weighted, respectively, by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.

@@

This description suggests that B\&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the \emph{exact same training data}, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

@@

B\&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war \emph{onset}, they are actually scored using \emph{incidence} of civil war. By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict. When forecasting rare events, it is common to have many false positives, i.e.~instances where we predict onset but no onsets occur. \emph{Ex ante}, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).

@@

This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results. Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves \emph{prior} to AUC calculation.

ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B\&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

@@

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)? Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to ``PITF Only'', smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average. The Escalation model, however, \emph{always} benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B\&S report, namely that the escalation model is generally superior.

@@

Adding PITF predictors actually improves performance in both cases; the ``With PITF Predictions'' model strictly dominates the ``Escalation Only'' model by quite a margin in the 6-month data version.

@@

The weighted model performs roughly on par with the Escalation Only model. One finding that remains is that the ``PITF Only'' model is outperformed by the ``Escalation Only'' model. As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor especially noteworthy.

@@

B\&S forecast the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assessed the accuracy of these forecasts. The forecasts themselves are probabilistic, i.e.~range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts (``1'') and the remaining forecasts to have been for no onset (``0''). They evaluate two alternative codings of observed civil war outcomes, one in which 15 countries, including Colombia, had ongoing civil wars at the end of 2015 (labeled ``Assuming Persistence''), and an alternative coding in which the civil war in Colombia ends but Turkey and Benin experience civil war onsets (``Assuming Change'').\footnote{\label{fn:leadDV}The former coding is in the main replication data, while the latter is hand-coded in the Stata script that produces their confusion matrices. There is also some potential ambiguity about the discussion in this section because the forecasts B\&S claim are for the first half of 2016 are actually for the second half of 2016, but scored against outcomes for the first half of 2016.} However, when they score the forecasts to produce the confusion matrices they report in their Table 4, they score them using civil war incidence, not onset. In fact, this a conscious choice B\&S seem to acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.

For reference for the following discussion, we reproduce confusion matrices for the 2016-H1 forecasts in Table \ref{tab:table4}, but distinguish between civil war onsets (\(Y=1\)), non-events (\(Y=0\)) that correspond to countries with an ongoing civil war, and non-events (\(Y=0\)) that correspond to countries without a civil war, but which could potentially experience a civil war onset. The original B\&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group. Essentially all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are actually ongoing civil war cases which by definition \emph{cannot} experience another civil war onset. The alternative outcome coding B\&S evaluate, under ``Assuming Change'', produces the only two civil war onsets we would have observed in 2016-H1.

@@

The fact that countries with an ongoing civil war cannot experience another civil war onset follows from the way the Sambanis civil war data conceptualize and operationalize civil war. Civil war appears to be a binary property of a country at any given moment in time, i.e.~a country at any given point is either at civil war or not, and onset is the first time period corresponding to this state. In the data, ongoing civil wars are coded as having a missing outcome value, and dropped.

If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that there are 15 countries that had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period. In all previous analyses in the paper, these kinds of cases drop out of the data. However, B\&S re-introduce, produce forecasts for them, and include them as positive events (``1''s) in their confusion matrices. They write, on page 21:

@@

It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense. A
conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, \emph{we would have already known that}. Rather than not making any ``assumptions about the persistence of civil war'', this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2. Since Sambanis is the author of the civil war data that is used, it would already have been known in early 2016 that these 15 civil wars had in fact not terminated, and were still ongoing. It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.

@@

B\&S advocate the use of theory to guide prediction. But ``theory'' is not defined by B\&S and may be an ambiguous and undefined concept. ``Theory'' has many interpretations in social science. For some---the EITM movement, for example (Aldrich, Alt, and Lupia 2008)---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced. The additional statements can be subjected to empirical scrutiny. An example might be William Riker's theory of coalitions (Riker 1962) which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example of strong theory is found in \emph{The Political Logic of Survival} (Bueno de Mesquita et al. 2005). For the purposes of discussion we call this strong theory.

@@

Furthermore, if what we mean by the concept ``theory'' is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model? We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off? Certainly that's not what B\&S are trying to suggest since this would be a problematic way to ``evaluate'' a theory. But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could in fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated ``theory'' with any eye towards out-of-sample predictive accuracy (cf.~Colaresi and Mahmood 2017 for how to avoid this problem while leveraging out-of-sample predictions). Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize and whose underlying theories may by extension not be externally valid (for a more detailed treatment of this issue, see Fariss and Jones 2018).

Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. It is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al. 2010), irregular leadership changes (Beger et al. 2017), and coups (Beger and Ward 2020).

Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are \emph{the} correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B\&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory (see also Fariss and Jones 2018)?

@@

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning. Even if B\&S were correct that in their study the theory based escalation model performed better, it would simply be a case study of a single case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory. Nor, against it.

@@

We encountered several issues in the code underlying the B\&S analysis. The problems we encountered are not subjective modeling choices. They bear no implications on the question of whether theory based model are preferable. However, when we fix these issues and perform an updated analysis, the B\&S conclusions are all essentially overturned. In other words, B\&S findings are based on a faulty analysis, invalid, and should be disregarded as evidence in favor of theory-based models.

>>>>

\begin{document}
\maketitle
\begin{abstract}
We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors. We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature. Fixing these mistakes results in a reversal of their claim that theory-based models of escalation are better at predicting onsets of civil war than other kinds of models. Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine. More importantly, we argue that rather than trying to contrast the roles of theory and "atheoretical" machine learning in predictive modeling, it would be more productive to focus on ways in which predictive modeling and machine learning could be used to strengthen extant predictive work. Instead, we argue that predictive modeling and machine learning are effective tools for theory testing.
\end{abstract}

@@

Although their analysis in the end does not support this point, we do not disagree that theory can be important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with ``theoretical'' modeling, B\&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms. They do not provide much guidance on an exact specification of what variables and which functional form is required. As a result, it is both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to ``explain'' a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

To examine the first question, B\&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analyses reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

\begin{itemize}
\tightlist
\item
  Escalation: a set of ten indicators, drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other.
\item
  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels. These are directed, thus making for four directed dyads, which with four quad categories make sixteen covariates.
\item
  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a total of four covariates.
\item
  CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling \(1,160\) covariates, which are mostly zero for any country in any month.
\item
  Average: an unweighted average of the predictions from the four models briefly described above.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Escalation Only: the original basic escalation model with only ICEWS predictors.
\item
  With PITF Predictors: a random forest that also adds the PITF annual, structural variables.
\item
  Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions.
\item
  PITF Split Population: the training data are split into high- and low-risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
\item
  PITF Only: a random forest model based only on the annual, structural PITF model predictors.
\end{itemize}

@@

Finally, B\&S used their escalation model to create forecasts for the first half of 2016. In their third analysis, they score the forecasts' accuracy using subsequent civil war onset data. This is summarized in B\&S Table 3.

@@

While replicating and analyzing B\&S's results, we found several issues worthy of further discussion and investigation. These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.\footnote{We also note additional concerns arising from the question of how the random forest models were tuned by B\&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.}

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model \emph{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B\&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. B\&S refer to smoothing only in the context of the ROC curves in Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90\%--17 of 19--used empirical ROC curves, and 2 (10\%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey. Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard. What is surprising is that anyone would choose the smoothed ROC implementation.

@@

We believe that B\&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weigh the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.

As a result: (1) in the test set, all countries with names that alphabetically start after ``Tongo'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical. For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions weighted by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.

@@

This description suggests that B\&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the \emph{same training data}, which consists of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

@@

B\&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war \emph{onset}, they are scored using \emph{incidence} of civil war. By definition, the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict. When forecasting rare events, it is common to have many false positives, i.e.~instances where we predict onset but no onsets occur. \emph{Ex ante}, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).

@@

This suggests that using smoothed ROC curves will have no substantive impact on the results; however, we find that smoothing does determine the comparative results. Their statement also implies that the ROC curves were only smoothed in referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves before AUC calculation.

ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result, they appear step-like. Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B\&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

@@

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)? Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to ``PITF Only'', smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average. The Escalation model, however, \emph{always} benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models and accounts for the incorrect result B\&S report, namely that the escalation model is generally superior.

@@

Adding PITF predictors improves performance in both cases; the ``With PITF Predictions'' model strictly dominates the ``Escalation Only'' model by quite a margin in the 6-month data version.

@@

The weighted model performs roughly on par with the Escalation Only model. One finding that remains is that the ``PITF Only'' model is outperformed by the ``Escalation Only'' model. As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising nor especially noteworthy.

@@

B\&S forecast the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assessed the accuracy of these forecasts. The forecasts themselves are probabilistic, i.e.~range from 0 to 1, but to score the forecasts they consider the 30 highest predictions to have been positive forecasts (``1'') and the remaining forecasts to have been for no onset (``0''). They evaluate two alternative codings of observed civil war outcomes, one in which 15 countries, including Colombia, had ongoing civil wars at the end of 2015 (labeled ``Assuming Persistence''), and an alternative coding in which the civil war in Colombia ends but Turkey and Benin experience civil war onsets (``Assuming Change'').\footnote{\label{fn:leadDV}The former coding is in the main replication data, while the latter is hand-coded in the Stata script that produces their confusion matrices. There is also some potential ambiguity about the discussion in this section because the forecasts B\&S claim are for the first half of 2016 are actually for the second half of 2016, but scored against outcomes for the first half of 2016.} However, when they score the forecasts to produce the confusion matrices they report in their Table 4, they score them using civil war incidence, not onset. This a conscious choice B\&S acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.

For reference for the following discussion, we reproduce confusion matrices for the 2016-H1 forecasts in Table \ref{tab:table4}, but distinguish between civil war onsets (\(Y=1\)), non-events (\(Y=0\)) that correspond to countries with an ongoing civil war, and non-events (\(Y=0\)) that correspond to countries without a civil war, but which could potentially experience a civil war onset. The original B\&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group. Essentially all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are ongoing civil war cases which by definition \emph{cannot} experience another civil war onset. The alternative outcome coding B\&S evaluate, under ``Assuming Change'', produces the only two civil war onsets we would have observed in 2016-H1.

@@

The fact that countries with an ongoing civil war cannot experience another civil war onset follows from the way the Sambanis civil war data conceptualize and operationalize civil war. Civil war appears to be a binary property of a country at any given moment in time, i.e.~a country at any given point is either at civil war or not, and onset is the first time period corresponding to this state. In the data, ongoing civil wars are coded as having a missing outcome value and dropped.

If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that 15 countries had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period. In all previous analyses in the paper, these kinds of cases drop out of the data. However, B\&S re-introduce, produce forecasts for them, and include them as positive events (``1''s) in their confusion matrices. They write, on page 21:

@@

It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense. A
conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, \emph{we would have already known that}. Rather than not making any ``assumptions about the persistence of civil war'', this would only make sense if we assume that all ongoing civil wars ended in 2015-H2. Since Sambanis is the author of the civil war data that is used, it would already have been known in early 2016 that these 15 civil wars had not terminated, and were still ongoing. It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.

@@

B\&S advocate the use of theory to guide prediction. But ``theory'' is not defined by B\&S and may be an ambiguous and undefined concept. ``Theory'' has many interpretations in social science. For some---the EITM movement, for example (Aldrich, Alt, and Lupia 2008)---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced. The additional statements can be subjected to empirical scrutiny. An example might be William Riker's theory of coalitions (Riker 1962) which posits propositions in game-theoretic form from which the size principle of minimal winning coalitions may be deduced and further examined empirically. A more recent example of strong theory is found in \emph{The Political Logic of Survival} (Bueno de Mesquita et al. 2005). For the purposes of discussion, we call this strong theory.

@@

Furthermore, if what we mean by the concept ``theory'' is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model? We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off? Certainly, that's not what B\&S are trying to suggest since this would be a problematic way to ``evaluate'' a theory. But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could, in fact, be tied to a theory, it is hard to demonstrate that this theory is not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated ``theory'' with an eye towards out-of-sample predictive accuracy (cf.~Colaresi and Mahmood 2017 for how to avoid this problem while leveraging out-of-sample predictions). Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize and whose underlying theories may by extension not be externally valid (for a more detailed treatment of this issue, see Fariss and Jones 2018).

Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for actually predicting those outcomes. It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al. 2010), irregular leadership changes (Beger et al. 2017), and coups (Beger and Ward 2020).

Aside from the selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility regarding functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are \emph{the} correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B\&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory (see also Fariss and Jones 2018)?

@@

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning. Even if B\&S were correct that in their study the theory-based escalation model performed better, it would simply be a case study of a single case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory. Nor, against it.

@@

We encountered several issues in the code underlying the B\&S analysis. The problems we encountered are not subjective modeling choices. They bear no implications on the question of whether theory-based model are preferable. However, when we fix these issues and perform an updated analysis, the B\&S conclusions are all essentially overturned. In other words, B\&S findings are based on a faulty analysis, invalid, and should be disregarded as evidence in favor of theory-based models.
